A simplified Plackett-Luce likelihood function for rank orders

### 2024-04-15 inst/hankin-sts.tex
repurpose for Statistical Science (STS)

\begin{abstract}
Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Placket-Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from the fields of Olympic
athletics, education, and Formula~1 motor racing.
\end{abstract}

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\cite{zipf1949} [a typical assumption for Bradley-Terry
  strengths], Hankin~\cite{hankin2017_rmd} shows that maximum
likelihood strengths identify the correct ranking with a probability
of essentially zero.  The situation is worse in cases such as formula
1 motor racing considered below, which requires us to consider
${\mathcal O}(100)$ competitors.

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$.  Edwards's two units of
support criterion~\cite{edwards1972} is met, and we are justified in
concluding that the student is indeed more competent at applied than
pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

>>>>

\begin{abstract}
Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett-Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from the fields of Olympic
athletics, education, and Formula~1 motor racing.
\end{abstract}

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\cite{zipf1949} [a typical assumption for Bradley-Terry
  strengths] and say 20 independent complete rank observations,
Hankin~\cite{hankin2017_rmd} shows that maximum likelihood strengths
identify the correct ranking with a probability of essentially zero.
The situation is worse in cases such as formula 1 motor racing
considered below, which requires us to consider ${\mathcal O}(100)$
competitors.

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$, at
${p}_\mathrm{pure}=0.567={p}_\mathrm{applied}=0.713$.  Edwards's two
units of support criterion~\cite{edwards1972} is met, and we are
justified in concluding that the student is indeed more competent at
applied than pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.
